Systematic Misestimation of Machine Learning Performance in Neuroimaging Studies of Depression

### 2021-05-03 ms.tex
fixed typos

%! language = Latex
\begin{document}
    \maketitle
    \begin{abstract}
        We currently observe a disconcerting phenomenon in machine learning studies in psychiatry: While we would expect larger samples to yield better results due to the availability of more data, larger machine learning studies consistently show much weaker performance than the numerous small-scale studies. Here, we systematically investigated this effect focusing on one of the mostheavily studied questions in the field, namely the classification of patients suffering from \ac{mdd} and \ac{hc} based on neuroimaging data. Drawing upon structural \ac{mri} data from a balanced sample of $N$\,=\;\num{1868} \ac{mdd} patients and \ac{hc} from our recent international \ac{pac}, we first trained and tested a classification model on the full dataset which yielded an accuracy of \SI{61}{\percent}. Next, we mimicked the process by which researchers would drawsamples of various sizes ($N$\,=\;\num{4} to $N$\,=\;\num{150}) from the population and showed a strong risk of misestimation. Specifically, for smallsample sizes ($N$\,=\;\num{20}), we observe accuracies of up to \SI{95}{\percent}. For medium sample sizes ($N$\,=\;\num{100}) accuracies up to \SI{75}{\percent} were found. Importantly, further investigation showed that sufficiently large test sets effectively protect against performance misestimationwhereas larger datasets per se do not. While these results question the validity of a substantial part of the current literature, weoutline the relatively low-cost remedy of larger test sets, which is readily available in most cases.

@@

    \begin{figure}
        \includegraphics[width=\textwidth]{./images/figure_1}
        \caption{Workflow to investigate the correlations between sample size and misestimation. First the effect of misestimation is investigated over the whole classification process ((1) Overall sample size analysis). Following the process of training and testing is evaluated seperatly ((2) Training set sample size analysis, (3) Test set sample size analysis)}
        \label{fig:workflow}
    \end{figure}

@@

    Given the profound effect of the test set size on systematic misestimation, it is important to consider why an effect of overestimation may arise. Previous work by Schnack \& Kahn~\cite{Schnack2016} suggests that patient characteristics in smaller samples tend to be more homogenous. In the case of small $N$, participants may be more likely to be recruited from the same data collection site and of a similar age (for example, in the case of a university recruited convenience sample). In addition, stringent recruitment criterion may be easily met, resulting in a well-defined phenotype that is not truly representative of the parent population of interest. Whilst this explanation makes sense for samples collected in this manner, it fails to explain why we observed this phenomenon in our random sampling procedure, and more importantly, with our dummy classifiers that paid no attention to participants, their characteristics, or the inputted predictor variables. This observation suggests a mechanism for systematic misestimation that is not just sample/patient-specific or contingent on sample homogeneity, but instead, inherent in the natural variation that arises in small test samples. Indeed, this effect is known as sampling error, and as demonstrated by Combrisson et al.~\cite{Combrisson2015} can lead to an effect whereby we exceed a machine learning model's chance level, purely by chance.

>>>>

%! language = Latex
\begin{document}
    \maketitle
    \begin{abstract}
        We currently observe a disconcerting phenomenon in machine learning studies in psychiatry: While we would expect larger samples to yield better results due to the availability of more data, larger machine learning studies consistently show much weaker performance than the numerous small-scale studies. Here, we systematically investigated this effect focusing on one of the most heavily studied questions in the field, namely the classification of patients suffering from \ac{mdd} and \ac{hc} based on neuroimaging data. Drawing upon structural \ac{mri} data from a balanced sample of $N$\,=\;\num{1868} \ac{mdd} patients and \ac{hc} from our recent international \ac{pac}, we first trained and tested a classification model on the full dataset which yielded an accuracy of \SI{61}{\percent}. Next, we mimicked the process by which researchers would draw samples of various sizes ($N$\,=\;\num{4} to $N$\,=\;\num{150}) from the population and showed a strong risk of misestimation. Specifically, for smallsample sizes ($N$\,=\;\num{20}), we observe accuracies of up to \SI{95}{\percent}. For medium sample sizes ($N$\,=\;\num{100}) accuracies up to \SI{75}{\percent} were found. Importantly, further investigation showed that sufficiently large test sets effectively protect against performance misestimation whereas larger datasets per se do not. While these results question the validity of a substantial part of the current literature, we outline the relatively low-cost remedy of larger test sets, which is readily available in most cases.

@@

    \begin{figure}
        \includegraphics[width=\textwidth]{./images/figure_1}
        \caption{Workflow to investigate the correlations between sample size and misestimation. First the effect of misestimation is investigated over the whole classification process ((1) Overall sample size analysis). Following the process of training and testing is evaluated separately ((2) Training set sample size analysis, (3) Test set sample size analysis)}
        \label{fig:workflow}
    \end{figure}

@@

    Given the profound effect of the test set size on systematic misestimation, it is important to consider why an effect of overestimation may arise. Previous work by Schnack \& Kahn~\cite{Schnack2016} suggests that patient characteristics in smaller samples tend to be more homogeneous. In the case of small $N$, participants may be more likely to be recruited from the same data collection site and of a similar age (for example, in the case of a university recruited convenience sample). In addition, stringent recruitment criterion may be easily met, resulting in a well-defined phenotype that is not truly representative of the parent population of interest. Whilst this explanation makes sense for samples collected in this manner, it fails to explain why we observed this phenomenon in our random sampling procedure, and more importantly, with our dummy classifiers that paid no attention to participants, their characteristics, or the inputted predictor variables. This observation suggests a mechanism for systematic misestimation that is not just sample/patient-specific or contingent on sample homogeneity, but instead, inherent in the natural variation that arises in small test samples. Indeed, this effect is known as sampling error, and as demonstrated by Combrisson et al.~\cite{Combrisson2015} can lead to an effect whereby we exceed a machine learning model's chance level, purely by chance.
